---
title: "Reproducible Research Project 1"
author: "Edem Dossou"
date: "May 7, 2016"
output: html_document
---


## Loading the data

```{r, echo=TRUE}
pacman::p_load(dplyr, data.table, lubridate, ggplot2, plotly,dygraphs)

#read in the data
dt <- fread("activity.csv")
dt$date <- ymd(dt$date)
#Let's explore the data by checking a quick summary, its structure and its dimensions
dims <- dim(dt) ; summary(dt); class(dt)
dims
```

## Mean total number of steps taken daily 

```{r, echo=TRUE}
#Total number of steps per day
steps.tot <- dt %>%
                     group_by(date) %>%
                     summarize(steps = sum(steps)) %>% 
                     na.omit 
steps.tot 

#Histogram of the total number of steps taken per day
p <- ggplot(data = steps.tot, aes(x=steps)) + 
        geom_histogram(binwidth = 2000, fill ="#0072B2") + 
        xlab("Daily number of steps") + ylab("Daily Frequency") + labs(title= "Total numbers of Steps per day") +
        theme_bw()
p <- ggplotly(p)
p

# Mean and median of the total number of steps taken per day summarized in a table
mm.dt <- data.table(mean.steps = mean(steps.tot$steps), median.steps = median(steps.tot$steps))
mm.dt
```

## Average daily activity pattern

```{r, echo=TRUE}
#get the time series data
ts.steps <- dt %>% 
                  filter(!is.na(steps)) %>%
                  group_by(interval) %>% 
                  summarize(steps = sum(steps))
ts.steps

#plot the time-series
dygraph(ts.steps) %>% dyRangeSelector()

#Interval with the maximum number of steps
int.val <- ts.steps[which.max(ts.steps$steps), ]$interval
int.val
```

## Imputing missing values

```{r, echo=TRUE}
#report the number of missing values in the data
tot.missing.vals <- sum(is.na(dt))
#Strategy for imputing the dataset - replacing missing values with meadian (steps and intervals)
imput.dt <- dt %>%
                 group_by(interval) %>%
                 mutate(steps = ifelse(is.na(steps),median(steps, na.rm = TRUE), steps))
summary(imput.dt)

#Total number of steps per day
steps.tot1 <- imput.dt %>%
        group_by(date) %>%
        summarize(steps = sum(steps))
steps.tot1 

#Histogram of the total number of steps taken per day               
g <- ggplot(data = steps.tot1, aes(x=steps)) + 
        geom_histogram(binwidth = 2000, fill ="#CC79A7") + 
        xlab("Daily number of steps") + ylab("Daily Frequency") + labs(title= "Total numbers of Steps per day") +
        theme_bw()
g <- ggplotly(g)
g

# Mean and median of the total number of steps taken per day summarized in a table
mm.dt2 <- data.table(mean.steps = mean(steps.tot1$steps), median.steps = median(steps.tot1$steps))
mm.dt2

#Function to return differences in the original and imputed steps 
sum_diff <- function(x,y) {
        diff.int <- summary(x) - summary(y)
        diff.bool <- all.equal(summary(x), summary(y))
        result <- list(difference = diff.int, equality = diff.bool)
        
        return(result)
}
```

There has been a 9.5% increase in the steps value given my strategy - filling missing steps with the median of the steps

```{r, echo=TRUE}
sum_diff(steps.tot$steps, steps.tot1$steps)
```

## Comparing weekdays and weekends activity pattern

Creating a factor varaible to convert with values the days of the week given a date and make panel plots for weekdays and weekends

```{r, echo=TRUE}
#creating a factor varaible to convert with values the days of the week given a date
imput.dt$days <- weekdays(imput.dt$date)
imput.dt <- imput.dt %>% 
            mutate(days_bin=ifelse(days %in% c("Friday","Monday","Thursday","Tuesday","Wednesday"), "weekday", "weekend"))
imput.dt$days_bin <- factor(imput.dt$days_bin)

#Make panel plots for weekdays and weekends
ts.steps2 <- imput.dt %>% 
        filter(days_bin %in% c("weekday","weekend")) %>%
        group_by(days_bin, interval) %>% 
        summarize(steps = mean(steps))
ts.steps2

#plots      
h <- ggplot(ts.steps2, aes(x=interval, y=steps)) +
                geom_line(aes(y = steps, colour = "mean"), size = .5) + 
                scale_colour_manual(name='', values=c("mean"="turquoise")) +
                facet_wrap(~days_bin, nrow=2) + 
                theme_bw() + theme(legend.position="none")
h <- ggplotly(h)
h

```








